\thetitle

### 2008-09-21 paper/paper.tex
paper: Fix lots of things (layout, code formatting, code itself, wording). Truncate things. Change conclusion somewhat.  src: Simplify some things. Fix many small todo's. Add some stuff for reasoning about the simpler quicksort definitions, used by the correctness and equivalence proofs (to be added soon).
darcs-hash:20080921225058-bab43-f34d7b096a00b3d10121dae840ec9d423155ba05.gz

\usepackage{color}
\usepackage[colorlinks,pdftitle={dus},pdfauthor=\theauthors]{hyperref}

@@

@@

@@

%format M_NDP
%format le_NDP
%format qs_NDP

@@

% todo:
% - check that names are spelled and encoded right in bibliography
% - lhs2tex shows parameters declared with {...} differently from those declared with (...)
% - mention that we don't do correctness proofs (although we can add them in the formalization, of course)
% - decide whether to hide, ignore, or explain R/N differences. explaining costs precious space. ignoring is "begging the question". hiding is naughty.
% - have a sentence somewhere hinting at the complications caused by duplicate elements (e.g., by pointing out simplifications that hold only with the assumption of no duplicate elements).

@@

\newcommand{\todo}[1]{\textcolor{magenta}{[TODO: #1]}}

@@

\section{Introduction}

@@

The formalization is based on the ``paper proof'' in \cite{introtoalgos}, which consists of three parts. The first part argues that the total number of comparisons performed by the algorithm can be written as the sum of the expected number of comparisons between individual pairs of elements from the input list (ordered a certain way). The second part then derives from the algorithm a specific formula for this expectancy. Finally, the third and last part employs some analysis involving the harmonic series to derive the $O(n \log n)$ bound from the sum-of-expectancies. Thus, the first part essentially establishes a lemma about the expected number of comparisons between individual pairs of elements from the input list as an interface between the second and third part, proved by the former and used by the latter.

Of these three parts, only the first two require reasoning about the actual behavior of the algorithm---the third part is just ordinary mathematics. While the original proof provides a thorough treatment of the third part, its treatment of the first two parts is informal in two major ways.

First, it never actually justifies anything in terms of the algorithm's formal semantics. Indeed, it does not even formally define the algorithm in the first place. Instead, it relies on assertions which are taken to be intuitively true. While this practice is common and perfectly reasonable for paper proofs intended for human consumption, it is a luxury we do not afford ourselves.

Second, the original proof (implicitly) assumes that the input list does not contain any duplicate elements, which significantly simplifies its derivation of the formula for the probability of any two given elements being compared by the algorithm. We do not make this assumption.
  % todo: bring this up later.

The key to giving a proper formal treatment of both of these issues lies in the use of an appropriate representation of the algorithm, capable of capturing its computational behavior---specifically, its use of comparisons---in a way suitable for subsequent formal reasoning. The representation we use takes the form of a shallow monadic embedding, in which the algorithm is written as a monadically expressed functional program. This definition is then instantiated with refinements of operation-counting monads to make the number of comparisons performed observable.

The embedding is introduced in section \ref{embed}, where we demonstrate its use by first defining a simple deterministic Quicksort implementation (which just selects the head of the input list as its pivot element, and uses two separate filter passes to partition the input list), and then defining an operation counting monad that lets us prove that implementation's quadratic worst-case complexity.

In preparation of the more complex average-case theorem, we then define (in section \ref{nondetexpec}) a nondeterministic Quicksort implementation which selects pivot elements at random, and which uses a single three-way partition pass instead of a two two-way filter passes. We also introduce a monad that combines operation counting with nondeterminism, supporting a formal definition of the notion of the \emph{expected} number of comparisons that are performed.

Instantiating the nondeterministic implementation with this monad then lets us state the main theorem in section \ref{statement}.

The next two sections detail the actual formalized proof. Section \ref{reduction} corresponds to the first part in the original proof, and shows how the main theorem can be split up into a lemma (stated in terms of another special-purpose monad) about the expected number of comparisons being performed for pairs of input elements indexed a certain way, and a strictly numerical part.

Since we were able to fairly directly transcribe the latter part from the paper proof, using the existing real-number theory in the Coq standard library with few complications and additions, we omit a discussion of it in this paper and refer the interested reader to the paper proof.

Section \ref{finishing} shows the proof of the lemma mentioned above, and corresponds to the part in the original proof that derives the expectancy formula from the algorithm. Since this is the part where the original proof omits the most detail, and makes the assumption regarding duplicate elements, and where we really have to reason in detail about the behavior of the algorithm, it is by far the most involved part of the formalization.

@@

% todo: less pompous here \/

As stated before, the key to giving a proper formal treatment of those parts of the proof for which the original contents itself with appeals to intuition, lies in the use of an appropriate representation of the algorithm. Indeed, we cannot even formally state the main theorem until we have both an algorithm definition and the means to denote its use of comparisons (the usual complexity metric for sorting algorithms).

@@

  Fixpoint filter (c: T -> M bool) (l: list T): M (list T) :=
    match l with
    | nil => ret nil
    | h :: t =>
      t' <- filter c t;
      b <- c h;
      ret (if b then h :: t' else t')
    end.

  Program Fixpoint qs (l: list T) {measure length l}: M (list T) :=
    match l with
    | nil => ret nil
    | pivot :: t =>
        lower <- filter (gt pivot) t >>= qs;
        upper <- filter (le pivot) t >>= qs;
        ret (lower ++ pivot :: upper)
    end.
\end{code}
|Monad| is a dependent record containing the (coercible) carrier of type |Set -> Set|, along with |bind| (infix: |>>=|) and |ret| (for ``return'') operations, and the three monad laws. For a general introduction to monadic programming and monad laws, see \cite{wadler93monads}. The notation ``|x <- y; z|'', shorthand for |y >>= \x => z| (where |x| may occur in |z|) is known as ``do-notation''.

@@

By instantiating the above definitions with the right monad, we can transparently insert comparison-counting instrumentation into the algorithm, which will prove to be sufficient to let us reason about its complexity. But before we do so, let us note that if the above definitions are instead instantiated with the identity monad and an ordinary elementwise comparison on |T|, then the monadic scaffolding melts away, and the result is equivalent to an ordinary non-instrumented, non-monadic version, suitable for extraction and correctness proofs. This means that while we will instantiate the definitions with less trivial monads to support our complexity proofs, we can take some comfort in knowing that the object of those proofs is, in a very concrete sense, the actual Quicksort algorithm (as one would write it in a functional programming language), rather than some idealized model thereof.

For reasons that will become clear in later sections, we compose the monad we will instantiate the above definitions with using a monad transformer called |MMT| (for ``monoid monad transformer''), which piggybacks a monoid onto an existing monad by pairing.

@@

  Eval compute in qs_SP (3 :: 1 :: 0 :: 4 :: 5 :: 2 :: nil).
    = (16, 0 :: 1 :: 2 :: 3 :: 4 :: 5 :: nil)
\end{code}
Defining |cost| and |result| as the first and second projection, respectively, we have the identities
\begin{code}
  forall x, cost (ret_SP x) = 0,
  forall x f, cost (bind_SP x f) = cost x + cost (f (result x),
  forall x y, cost (le_SP x y) = 1.
\end{code}
This very modest amount of machinery is sufficient for a straightforward proof of Quicksort's quadratic worst-case complexity.

\begin{propo}|qs_worst: forall l, cost (qs_SP l) <= sqrd (length l).|\footnote{(We don't use big-$O$ notation for this simple statement, as it would only obfuscate. Big-$O$ complexity is discussed in section \ref{statement}.)}
\end{propo}

\begin{proof}
  The proof is by induction matching |qs|'s recursion. For an empty input list, we have |cost (qs_SP nil) = cost (ret nil) = 0 <= sqrd (length l)|. For a non-empty input list |(pivot :: t)|, the cost decomposes into
  \begin{code}
    cost (filter (le pivot) t) + cost (qs_SP (result (filter (lt pivot) t))) +
    cost (filter (gt pivot) t) + cost (qs_SP (result (filter (gt pivot) t))) +
    cost (ret (result (qs_SP (result (filter (lt pivot) t))) ++
     pivot :: result (qs_SP (result (filter (gt pivot) t))))).
  \end{code}
  The |filter| costs are easily proved (by induction on |t|) to be |length t| each. The cost of the final |ret| is 0 by definition. The induction hypothesis applies to the recursive |qs_SP| calls. Furthermore, by induction on |t|, we can easily prove
  \begin{code}
  length (result (filter (le pivot) t)) +
  length (result (filter (gt pivot) t)) <= length t,
  \end{code}
  because the two predicates filtered on are mutually exclusive. Abstracting |filter (le pivot) t| and its |gt| counterpart, this leaves
  \begin{code}
    forall (t flt flt': list T), length flt + length flt' <= length t ->
      length t + sqrd (length flt) + length t + sqrd (length flt') + 0 <= sqrd (S (length t)),
  \end{code}
  which is true by elementary arithmetic. \qed
\end{proof}

@@

The Quicksort algorithm used by the average-case proof we formalize differs from the one presented in the last section in two ways. First, it selects a pivot element nondeterministically, thereby avoiding the pathological quadratic cases any deterministic pivot selection strategy inevitably suffers. Second, it uses a single three-way partition pass instead of two two-way filter passes, to avoid the pathological quadratic case that two two-way filter passes suffer for an input list consisting of elements all of the same value.

Combined, these two traits ensure that the $O(n \log n)$ average-case bound holds not just averaged over all input lists, but for each individual input list as well. This greatly simplifies things, because it means that the global bound can be proved simply by proving it as a bound of the expected number of comparisons performed for an arbitrary input.

Hence, we now modify our |qs| definition to use a three-way partition pass, and to use nondeterministic pivot selection. Following that, we will compose a new monad (again using |MMT|, but this time transforming a nondeterminism monad) that the new definition can be instantiated with to capture its \emph{expected} number of comparisons.

First, we define |partition|. Instead of a two-way comparison operation producing a monadic |bool|, it takes a three-way comparison operation producing a monadic |comparison|, which is an enumeration with values |Lt|, |Eq|, and |Gt|. |partition| constructs a |Partitioning T|, which is simply a function of type |comparison -> list T|. We might equally well have used an inductive definition of |Partitioning T| as a record containing three lists. % at least until i remember why i used a function.

\parbox{\textwidth}{\begin{code}
  Variables (T: Set) (M: Monad) (cmp: T -> T -> M comparison).

  Fixpoint partition (pivot: T) (l: list T): M (Partitioning T) :=
    match l with
    | nil => ret emptyPartitioning
    | h :: t =>
        b <- cmp h pivot;
        tt <- partition pivot t ;
        ret (addToPartitioning b h tt)
    end.
\end{code}}

@@

\begin{code}
  Variable pick: forall (A: Set), ne_list A -> M A.

@@

Nondeterminism can now be achieved by instantiating these definitions with a suitable monad, along with a corresponding |pick| operation. A deterministic, non-instrumented version can still be obtained, simply by instantiating the definitions with the identity monad and a corresponding deterministic |pick| operation, like |head|.

Let us now consider what kind of nondeterminism monad would be suitable for reasoning about the expected value of a nondeterministic program. The list monad is commonly used to emulate nondeterministic computation. With the list monad, the program
\begin{code}
  x <- pick [0, 1]; if x = 0 then ret 0 else pick [1, 2],
\end{code}
produces |[0, 1, 2]| as its list of possible outcomes. Unfortunately, the information that 0 is a more likely outcome than 1 or 2, has been lost. Such relative probabilities are critical to the notion of an expected value: the expected value of the program above is |avg [0, avg [1, 2]] = {-"\frac{3}{4}"-} /= 1 = avg [0, 1, 2]|. This makes list nondeterminism unsuitable for our purposes.

@@

  Fixpoint bind_ne_tree (A B: Set) (m: ne_tree A) (k: A -> ne_tree B): ne_tree B :=
    match m with
    | Leaf a => k a
    | Node ts => Node (ne_list.map (fun x => bind_ne_tree x k) ts)
    end.

  Definition M_ne_tree: Monad := Build_Monad ne_tree bind_ne_tree ret_ne_tree.

@@

|expec| gives rise to a host of identities, including

@@

  Definition le_NDP (x y: T): M bool := ret_ne_tree (1, le x y).

  Definition qs_NDP := qs M_NDP le_NDP (lift pick_ne_tree).
\end{code}
We can now denote the expected number of comparisons of a |qs_NDP| application by |expec cost (qs_NDP l)|, and will use this in our statement of the main theorem in the next section.

But before we do so, we define a slight refinement of |expec| that specifically observes the monoid component of computations in monads formed by transforming |M_ne_tree| using |MMT| (like |NDP|).
\begin{code}
  Definition monoid_expec (m: Monoid) (f: m -> nat) {A: Set}
    : (MonoidMonadTrans.M m NeTreeMonad A) -> R := expec (f fcmp fst).
\end{code}
Since |cost = fst|, we have |forall t, expec cost t = monoid_expec id t|.

@@

\parbox{\textwidth}{\begin{code}
  monoid_expec_plus: forall (m: Monoid) (h: m -> nat) (A B: Set)
    (f: MMT m NeTreeMonad A) (g: A -> MMT m NeTreeMonad B):
    (forall x y `In` f -> monoid_expec h (g (snd x)) = monoid_expec h (g (snd y))),
    monoid_expec h (f >>= g) =
      monoid_expec h f + monoid_expec h (g (snd (ne_tree.head f))).
\end{code}}

@@

The last thing needed before the main theorem can be stated, is the notion of big-O complexity. The definition we use is a standard textbook definition, except that we work with a separate measure function parameter:

\parbox{\textwidth}{\begin{code}
  Definition measured_bigO (X: Set)
    (m: X -> nat) (f: X -> R) (g: nat -> R): Prop
      := exists c, exists n, forall x, n <= m x -> f x <= c * g (m x).

  Notation {-"``"-}over m ,{-"\ "-} f =O( g ){-"\text{''''}"-} := (measured_bigO m f g).
\end{code}}
We now state the main theorem.
\begin{code}
  Theorem qs_avg: over length,
    expec cost fcmp qs_NDP =O(fun n => INR (n * log2ceil n)).
\end{code}
Thanks to the property discussed at the start of the previous section, % perhaps too far back
the above follows as a corollary from the (slightly stronger) statement that
\begin{code}
  Theorem qs_expec_cost:
    forall l, expec cost (qs_NDP l) <= 2 * length l * S (log2ceil (length l)),
\end{code}
the proof of which is described in the remainder of this paper.

@@

The main idea in the proof is to reduce |qs_expec_cost| to a statement about the expected number of comparisons between specific pairs of elements from the input list.

If $X \equiv X_{I_0} \ldots X_{I_{n-1}}$ is the input list, with $I$ a permutation of |[0 ... n-1]| such that $X_0 \ldots X_n$ is sorted, then the expected number of comparisons between any $X_i$ and $X_j$ with $i < j$ is at most $2 / S (j - i)$. In other words, the expected number of comparisons between two list elements is bounded by a simple function of the number of list elements that separate the two in the sort order. We prove this fact in the next section, but first show how |qs_expec_cost| follows from it.

Combined with the observation that the total expected number of comparisons equals the sum of the expected numbers of comparisons for each |(i, j)| in $\Varid{IJ} := \{ (i, j) \in [0, |length l|) \mid i < j \}$, the property described above suggests breaking up the inequality into
\begin{code}
  expec cost (qs_NDP l) <= {-"\displaystyle{\sum_{(i, j) \in \Varid{IJ}}\frac{2}{S (j - i)}}"-} <= 2 * length l * S (log2ceil (length l)).
\end{code}
Proving the right inequality requires a bit of analysis involving the harmonic series. This part of the proof could be fairly directly transcribed from the paper proof, using the existing real-number theory in the Coq standard library, with few complications and additions. We refer the interested reader to the paper proof.

The left inequality is the challenging one. To bring it closer to the index summation, we first rewrite
\begin{code}
  expec cost (qs_NDP l) = expec cost (qs_NDP (map (nth (sort l)) li)),
\end{code}
where |sort| may be any sorting function (including |qs| itself), and where |li| is a permutation of |[0...n - 1]| such that |map (nth (sort l)) li = l| (such an |li| can easily be proven to exist).

@@

  Definition cmp_U (x y: nat): U comparison :=
    ret ((if x <= y then (x, y) else (y, x)) :: nil, lookup_cmp x y).

  Definition qs_U: list nat -> list nat := qs cmp_U pick_U.
\end{code}
|qs_U| operates directly on a list of natural numbers representing indices into |sort l|. Comparison of indices is defined by comparison of the |T| values they denote in |sort l|. Furthermore, rather than producing a grand total comparison count the way |NDP| does, |U| records every pair of compared indices, by using |MMT| with |Monoid_U|, the free monoid over |nat * nat| pairs, instead of the additive monoid on |nat| we used up until now.

The goal can now be rewritten using
\begin{code}expec cost (qs_NDP (map (nth (sort l)) li))
  = monoid_expec length (qs_U li) = expec (length fcmp fst) (qs_U li).
\end{code}
The first equality expresses that the expected number of comparisons counted by |NDP| is equal to the expected length of the list of comparisons recorded by |U|. In the code, it is proved as a separate lemma by induction matching either side's |qs|'s recursion. The second equality merely unfolds the definition of |monoid_expec|.

@@

\parbox{\textwidth}{\begin{code}
  forall (X: Set) (fr: X -> R) (q: list X) (t: ne_tree (list X)):
    (forall x `In` q, expec (count x) t <= fr i) ->
    (forall x `NotIn` q, expec (count x) t = 0) ->
    NoDuplicates q -> expec length t <= {-"\displaystyle{\sum_{\Varid{i} \in \Varid{q}}\Varid{fr}\ \Varid{i}}"-}.
\end{code}}
It is trivially the case that |IJ| does not contain duplicate values. The second subgoal generated is\begin{code}forall (i, j) `NotIn` IJ, expec (count (i, j)) (ne_tree.map fst (qs_U li)) = 0.\end{code} Rewriting this using  the last of the identities listed for |expec| in section \ref{nondetexpec} in reverse, then rewriting the |expec| as a |monoid_expec|, and then generalizing the premise, results in the following property which is useful in its own right, as we will see later.
\begin{code}sound_cmp_expec_0: forall i j li, (i `NotIn` li \/ j `NotIn` li) ->
  monoid_expec (count (i, j)) (U.qs li) = 0,\end{code}
The proof of |sound_cmp_expec_0|, which we won't show here, is by a separate induction matching |qs|'s recursion.

This leaves but one subgoal which, expressed with |monoid_expec|, reads
\begin{code}
  forall (i, j) `In` IJ, monoid_expec (count (i, j)) (qs_U li) <= 2 / INR (S (j - i)).
\end{code}
This is exactly  to the property described at the beginning of this section. We prove it in the next section.

@@

In the interest of brevity, we have omitted innumerable details and lemmas in the description of the proof. Still, the parts shown are reasonably faithful to the actual formalization. For completeness, we briefly list some complicating factors that we have ignored. In some cases, resolving these took significant effort.

>>>>

\usepackage{color}
\usepackage[colorlinks,pdftitle=\thetitle,pdfauthor=\theauthors]{hyperref}

@@

%format sp = "\ "

%format == = "= "
%format /= = "\ne "
  % Without these, lhs2tex uses three-line equal signs.

@@

%format { = "\hspace{0.9mm}\{"
  % so that parameters declared with {...} are spaced like those declared with (...)

@@

%format M_NDP
%format cmp_NDP
%format qs_NDP

@@

% todo:
% - check that names are spelled and encoded right in bibliography
% - decide whether to hide, ignore, or explain R/N differences. explaining costs precious space. ignoring is "begging the question". hiding is naughty.
% - have a sentence somewhere hinting at the complications caused by duplicate elements (e.g., by pointing out simplifications that hold only with the assumption of no duplicate elements).

@@

\arrayhs % i.e. no page breaks in code blocks

@@

\section{Introduction}
\label{intro}

@@

The formalization is based on the ``paper proof'' in \cite{introtoalgos}, which consists of three parts. The first part shows that the total number of comparisons performed by the algorithm (the usual complexity metric for sorting algorithms) can be written as a sum of expected comparison counts for individual pairs of input list elements. The second part derives from the algorithm a specific formula for this expectancy. The third and last part employs some analysis involving the harmonic series to derive the $O(n \log n)$ bound from the sum-of-expectancies.

Of these three parts, only the first two involve the actual algorithm itself---the third part is strictly numerical. While the original proof provides a thorough treatment of the third part, its treatment of the first two parts is informal in two major ways.

First, it never actually justifies anything in terms of the algorithm's formal semantics. Indeed, it does not even formally define the algorithm in the first place, relying instead on assertions which are taken to be intuitively true. While this practice is common and perfectly reasonable for paper proofs intended for human consumption, it is a luxury we do not afford ourselves.

Second, the original proof (implicitly) assumes that the input list does not contain any duplicate elements, which significantly simplifies its derivation of the formula for the expected comparison count for pairs of individual input list elements. We do not make this assumption.
  % todo: perhaps bring this up later.

The key to giving a proper formal treatment of both of these aspects lies in the use of an appropriate representation of the algorithm, capable of capturing its computational behavior---specifically, its use of comparisons---in a way suitable for subsequent formal reasoning. The representation we use takes the form of a shallow monadic embedding, in which the algorithm is written as a monadically expressed functional program. This definition is then instantiated with refinements of operation-counting monads to make the comparison count observable.

The embedding is introduced in section \ref{embed}, where we demonstrate its use by first giving a simple deterministic monadic Quicksort definition, and then instantiating it with a simple operation counting monad that lets us prove its quadratic worst-case complexity.

For the purposes of the more complex average-case theorem, we then give (in section \ref{nondetexpec}) a potentially-nondeterministic monadic Quicksort definition, and compose a monad that combines operation counting with nondeterminism, supporting a formal definition of the notion of the \emph{expected} comparison count, with which we state the main theorem in section \ref{statement}.

The next two sections detail the actual formalized proof. Section \ref{reduction} corresponds to the first part in the original proof described above, and shows how the main theorem can be split up into a lemma (stated in terms of another specialized monad) giving a formula for the expected comparison count for individual pairs of input element, and a strictly numerical part. Since we were able to fairly directly transcribe the latter part from the paper proof, using the existing real number theory in the Coq standard library with few complications and additions, we omit a discussion of it in this paper and refer the interested reader to the paper proof.

Section \ref{finishing} finishes the proof by proving the lemma about the expected comparison count for individual input list elements. Since this is the part where the original proof omits the most detail, and makes the assumption regarding duplicate elements, and where we really have to reason in detail about the behavior of the algorithm, it is by far the most involved part of the formalization.

@@

As stated before, the key to giving a proper formal treatment of those parts of the proof for which the original contents itself with appeals to intuition, lies in the use of an appropriate representation of the algorithm. Indeed, we cannot even formally state the main theorem until we have both an algorithm definition and the means to denote its use of comparisons.

@@

  Program Fixpoint qs (l: list T) {measure length l}: M (list T) :=
    match l with
    | nil => ret nil
    | pivot :: t =>
        lower <- filterM (gt pivot) t >>= qs;
        upper <- filterM (le pivot) t >>= qs;
        ret (lower ++ pivot :: upper)
    end.
\end{code}
|Monad| is a dependent record containing the (coercible) carrier of type |Set -> Set|, along with |bind| (infix: |>>=|) and |ret| (for ``return'') operations, and the three monad laws. For a general introduction to monadic programming and monad laws, see \cite{wadler93monads}. The notation ``|x <- y; z|'', shorthand for |y >>= \x => z| (where |x| may occur in |z|) is known as ``do-notation''.

@@

By instantiating the above definitions with the right monad, we can transparently insert comparison-counting instrumentation into the algorithm, which will prove to be sufficient to let us reason about its complexity. But before we do so, let us note that if the above definitions are instead instantiated with the identity monad and an ordinary elementwise comparison on |T|, then the monadic scaffolding melts away, and the result is equivalent to an ordinary non-instrumented, non-monadic version, suitable for extraction and correctness proofs (which are included in the formalization for completeness). This means that while we will instantiate the definitions with less trivial monads to support our complexity proofs, we can take some comfort in knowing that the object of those proofs is, in a very concrete sense, the actual Quicksort algorithm (as one would write it in a functional programming language), rather than some idealized model thereof.

For reasons that will become clear in later sections, we compose the monad we will instantiate the above definitions with using a monad transformer |MMT| (for ``monoid monad transformer''), which piggybacks a monoid onto an existing monad by pairing.

@@

  Eval compute in qs_SP (3 :: 1 :: 0 :: 4 :: 5 :: 2 :: nil).
    = (16, 0 :: 1 :: 2 :: 3 :: 4 :: 5 :: nil)
\end{code}

%format `infixbind_SP` = >>="_{SP}"

Defining |cost| and |result| as the first and second projection, respectively, we trivially have identities such as |cost (ret_SP x) = 0|, |cost (x `infixbind_SP` f) = cost x + cost (f (result x))|, and |cost (le_SP x y) = 1|. This very modest amount of machinery is sufficient for a straightforward proof of Quicksort's quadratic worst-case complexity.

\begin{propo}|qs_worst: forall l, cost (qs_SP l) <= sqrd (length l).|\footnote{(We do not use big-O notation for this simple statement, as it would only obfuscate. Big-O complexity is discussed in section \ref{statement}.)}
\end{propo}

\begin{proof}
  The proof is by induction matching |qs|'s recursion. For an empty input list, we have |cost (qs_SP nil) = cost (ret nil) = 0 <= sqrd (length l)|. For a non-empty input list |h :: t|, the cost decomposes into
  \begin{code}
    cost (filter (le h) t) + cost (qs_SP (result (filter (lt h) t))) +
    cost (filter (gt h) t) + cost (qs_SP (result (filter (gt h) t))) +
    cost (ret (result (qs_SP (result (filter (lt h) t))) ++
     h :: result (qs_SP (result (filter (gt h) t))))).
  \end{code}
  The |filter| costs are easily proved (by induction on |t|) to be |length t| each. The cost of the final |ret| is 0 by definition. The induction hypothesis applies to the recursive |qs_SP| calls. Furthermore, by induction on |t|, we can easily prove
  \begin{code}
  length (result (filter (le h) t)) + length (result (filter (gt h) t)) <= length t,
  \end{code}
  because the two predicates filtered on are mutually exclusive. Abstracting the filter terms as |flt| and |flt'|, this leaves
  \begin{code}
    length flt + length flt' <= length t ->
      length t + sqrd (length flt) + length t + sqrd (length flt') + 0 <= sqrd (S (length t)),
  \end{code}
  which is true by elementary arithmetic. \qed
\end{proof}

@@

The Quicksort algorithm used by the average-case proof we formalize differs from the one presented in the last section in two ways. First, it selects a pivot element nondeterministically, thereby avoiding the pathological quadratic cases any deterministic pivot selection strategy inevitably suffers. Second, it uses a single three-way partition pass instead of two two-way filter passes, thereby avoiding the pathological quadratic case that two two-way filter passes suffer for an input list whose elements are all of the same value.

Combined, these two traits ensure that the $O(n \log n)$ average-case bound holds not just averaged over all possible input lists, but for each individual input list as well. This greatly simplifies things, because it means that if we prove the bound for an arbitrary input, the global bound immediately follows.

Hence, we now modify our |qs| definition to use a three-way partition pass, and to use nondeterministic pivot selection. Following that, we will compose a new monad (again using |MMT|, but this time transforming a nondeterminism monad) that the new definition can be instantiated with to capture its \emph{expected} comparison count.

First, we define |partition|. Instead of a two-way comparison operation producing a monadic |bool|, it takes a three-way comparison operation producing a monadic |comparison|, which is an enumeration with values |Lt|, |Eq|, and |Gt|.

\begin{code}
  Variables (T: Set) (M: Monad) (cmp: T -> T -> M comparison).

  Fixpoint partition (t: T) (l: list T): M (comparison -> list T) :=
    match l with
    | nil => ret (const nil)
    | h :: l' =>
        c <- cmp h t; f <- partition t l';
        ret (fun c' => if c == c' then h :: f c' else f c')
    end.
\end{code}

@@

\begin{code}
  Variable pick: forall A: Set, ne_list A -> M A.

@@

Nondeterminism can now be emulated by instantiating these definitions with a suitable monad and |pick| operation. A deterministic, non-instrumented version can still be obtained, simply by using the identity monad and a deterministic |pick| operation, like |head|.

Let us now consider what kind of nondeterminism monad would be suitable for reasoning about the expected value of a nondeterministic program like
\begin{code}
  x <- pick [0, 1]; if x = 0 then ret 0 else pick [1, 2].
\end{code}
When executed in the list monad (commonly used to emulate nondeterministic computation), this program
produces |[0, 1, 2]| as its list of possible outcomes. Unfortunately, the information that 0 is a more likely outcome than 1 or 2, has been lost. Such relative probabilities are critical to the notion of an expected value: the expected value of the program above is |avg [0, avg [1, 2]] = {-"\frac{3}{4}"-} /= 1 = avg [0, 1, 2]|. This makes list nondeterminism unsuitable for our purposes.

@@

  Fixpoint bind_ne_tree (A B: Set) (m: ne_tree A) (k: A -> ne_tree B): ne_tree B :=
    match m with
    | Leaf a => k a
    | Node ts => Node (ne_list.map (fun x => bind_ne_tree x k) ts)
    end.
\end{code}
\begin{code}
  Definition M_ne_tree: Monad := Build_Monad ne_tree bind_ne_tree ret_ne_tree.

@@

|expec| gives rise to a host of identities, such as

@@

  Definition cmp_NDP (x y: T): M bool := ret_ne_tree (1, cmp x y).

  Definition qs_NDP := qs M_NDP cmp_NDP (lift pick_ne_tree).
\end{code}
We can now denote the expected comparison count for a |qs_NDP| application by |expec cost (qs_NDP l)|, and will use this in our statement of the main theorem in the next section.

But before we do so, we define a slight refinement of |expec| that specifically observes the monoid component of computations in monads formed by transforming |M_ne_tree| using |MMT| (like |NDP|).
\begin{code}
  Definition monoid_expec (m: Monoid) (f: m -> nat) {A: Set}
    : (MonoidMonadTrans.M m NeTreeMonad A) -> R := expec (f fcmp fst).
\end{code}
Since |cost = fst|, we have |expec cost t = monoid_expec id t|.

@@

\begin{code}
  monoid_expec_plus: forall (m: Monoid) (h: m -> NatAddMonoid),
    monoid_homo h -> forall (A B: Set)
    (f: MMT m NeTreeMonad A) (g: A -> MMT m NeTreeMonad B):
    (forall x y `In` f -> monoid_expec h (g (snd x)) = monoid_expec h (g (snd y))),
    monoid_expec h (f >>= g) =
      monoid_expec h f + monoid_expec h (g (snd (ne_tree.head f))).
\end{code}

@@

The last thing needed before the main theorem can be stated, is the notion of big-O complexity. We use a standard textbook definition, except that we incorporate a measure function parameter:

\begin{code}
  Definition bigO (X: Set) (m: X -> nat) (f: X -> R) (g: nat -> R): Prop
      := exists c n, forall x, n <= m x -> f x <= c * g (m x).

  Notation {-"``"-}over m,{-"\ "-} f =O( g ){-"\text{''''}"-} := bigO m f g.
\end{code}
We now state the main theorem.
\begin{code}
  Theorem qs_avg: over length, expec cost fcmp qs_NDP =O(fun n => INR (n * log2ceil n)).
\end{code}
Thanks to the property discussed at the start of the previous section, % perhaps too far back
|qs_avg| follows as a corollary from the stronger statement that
\begin{code}
  forall l, expec cost (qs_NDP l) <= 2 * length l * S (log2ceil (length l)),
\end{code}
the proof of which is described in the next two sections.

@@

As described in the introduction, the key ingredient in the proof is a lemma giving a formula for the expected comparison count for individual pairs of input list elements, indexed a certain way. More specifically, if $X \equiv X_{I_0} \ldots X_{I_{n-1}}$ is the input list, with $I$ a permutation of |[0 ... n-1]| such that $X_0 \ldots X_n$ is sorted, then the expected comparison count for any $X_i$ and $X_j$ with $i < j$ is at most $2 / S (j - i)$. In other words, the expected comparison count for two input list elements is bounded by a simple function of the number of list elements that separate the two in the sort order. We prove this fact in the next section, but first show how |qs_expec_cost| follows from it.

Combined with the observation that the total expected comparison count ought to equal the sum of the expected comparison count for each individual pair of input elements, the property described above suggests breaking up the inequality into
\begin{code}
  expec cost (qs_NDP l) <= {-"\displaystyle{\sum_{(i, j) \in \Varid{IJ}}\frac{2}{S (j - i)}}"-} <= 2 * length l * S (log2ceil (length l)),
\end{code}
where $\Varid{IJ} := \{ (i, j) \in [0, |length l|) \mid i < j \}$.

The right inequality is a strictly numerical affair, requiring a bit of analysis involving the harmonic series. As stated before, this part of the proof could be fairly directly transcribed from the paper proof, with few complications and additions, and so we will not discuss it.

The left inequality is the challenging one. To bring it closer to the index summation, we first write |l| on the left side as |map (nth (sort l)) li|, where |sort| may be any sorting function (including |qs| itself), and where |li| is a permutation of |[0...n-1]| such that |map (nth (sort l)) li = l| (such an |li| can easily be proven to exist).

@@

  Definition unordered_nat_pair (x y: nat): nat * nat :=
    if x <= y then (x, y) else (y, x).

  Definition cmp_U (x y: nat): U comparison :=
    ret (unordered_nat_pair x y :: nil, lookup_cmp x y).

  Definition qs_U: list nat -> list nat := qs U cmp_U pick_U.
\end{code}
|qs_U| operates directly on lists of indices into |sort l|. Comparison of indices is defined by comparison of the |T| values they denote in |sort l|. Furthermore, rather than producing a grand total comparison count the way |NDP| does, |U| records every pair of indices compared, by using |MMT| with |Monoid_U|, the free monoid over |nat * nat| pairs, instead of the additive monoid on |nat| we used up until now.

We now rewrite
\begin{code}expec cost (qs_NDP (map (nth (sort l)) li))
  = monoid_expec length (qs_U li) = expec (length fcmp fst) (qs_U li).
\end{code}
The first equality expresses that the expected number of comparisons counted by |NDP| is equal to the expected length of the list of comparisons recorded by |U|. In the formalization, this is a separate lemma proved by induction matching |qs|'s recursion. The second equality merely unfolds the definition of |monoid_expec|.

@@

\begin{code}
  forall (X: Set) (fr: X -> R) (q: list X) (t: ne_tree (list X)),
    (forall x `In` q, expec (count x) t <= fr x) ->
    (forall x `NotIn` q, expec (count x) t = 0) -> expec length t <= {-"\displaystyle{\sum_{\Varid{x} \in \Varid{q}}\Varid{fr}\ \Varid{x}}"-}.
\end{code}
We end up with two subgoals, the first of which is \begin{code}forall (i, j) `NotIn` IJ, expec (count (i, j)) (ne_tree.map fst (qs_U li)) = 0.\end{code} Rewriting this using  the last of the identities listed for |expec| in section \ref{nondetexpec} in reverse, then rewriting the |expec| as a |monoid_expec|, and then generalizing the premise, results in
\begin{code}
  forall i j li, (i `NotIn` li \/ j `NotIn` li) -> monoid_expec (count (i, j)) (U.qs li) = 0,
\end{code}
which can be shown by induction matching |qs|'s recursion, although we will not do so in this paper. We \emph{will} use this property again in the next section.

This second subgoal, expressed with |monoid_expec|, reads
\begin{code}
  forall (i, j) `In` IJ, monoid_expec (count (i, j)) (qs_U li) <= 2 / INR (S (j - i)).
\end{code}
This corresponds exactly to the property described at the beginning of this section. We prove it in the next section.

@@

In the interest of brevity, we have omitted lots of detail and various lemmas in the description of the proof. Still, the parts shown are reasonably faithful to the actual formalization. For completeness, we briefly list some complicating factors that we have ignored. In some cases, resolving these took significant effort.
